Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation

### 2022-11-09 article/abs.Rmd
bioRxiv,poster

<!-- ## REVIEW CRITERIA -->
<!-- ​ -->
<!-- All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee. The Scientific Committee may accept, decline, or offer an opportunity for a revised abstract submission. Proposals will be assessed using the following criteria:  -->

@@

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of previously degraded forests.
Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to broad dispersal mode and wood density traits using relevant literature.

<!-- # Results/Conclusions -->
As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

@@

# Resumen*{-}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

>>>>

<!-- ## REVIEW CRITERIA -->
<!-- . -->
<!-- All abstracts will be evaluated and scored by the ATBC2022 Scientific Committee. The Scientific Committee may accept, decline, or offer an opportunity for a revised abstract submission. Proposals will be assessed using the following criteria:  -->

@@

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of nearby degraded forests.
Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
A wet neotropical timber plantation reforested after 20 years and adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched identified taxa to broad dispersal mode and wood density traits using relevant literature.

<!-- # Results/Conclusions -->
As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light surprisingly tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, explained in part by wood density, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, as well as genus Ficus, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

@@

# Resumen{-}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, tanto como el género Ficus, pero biomasa por modo de disperso no tendió a cambiar.
Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

### 2022-11-09 article/disc.Rmd
bioRxiv,poster

This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
<!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
It was initially hypothesized that community composition would vary with distance to edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential, 
and the evidence presented here supported edge effects on community composition and biomass of taxa associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation *[@krishnadas20]*, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.

Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation *[@brinck17]*, with more expected in the future *[@mitchard18]*.
<!-- responds variably to secondary succession and edge effects ...   -->
Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
This study adds that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04a]* together with other global change drivers *[@laurance14]*, as well as may reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
    <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->

Edge effects have been shown in other studies to lower both forest biomass and change community composition *[@anderson22]*, and this study adds that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of lasting changes to community composition.
This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
Another possible factor may be overall short dispersal kernels, in part due to higher fungal pathogen pressure under higher constant shade, also increasing tree mortality.
Decomposing compositional changes into key specific taxa, while the biomass of the most abundant taxon *Vochysia* was resilient to edge effects in this study, that of the key taxon for frugivores like birds and bats *Ficus* was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and food web associated ecosystem services.
Community composition also tended to show lower variance across the edge gradient, although this effect cannot be confirmed here given the uneven sampling that accounted for variable forest area at different distances from edge.

Characterizing community composition by functional traits is also increasingly studied *[@kearney21]*, but less so regarding edge effects.
A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass *[@razafindratsima18]*.
These results parallel in showing lower taxonomic diversity, although biomass specifically of taxa assoociated with both early- and late-successional stages also decreased with away from primary forest edge.
This result by successional stage can be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge.
While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with dispersal kernels, the lack of pattern by broad dispersal mode limits this explanation here, yet could be further tested by measuring taxon-specific dispersal kernels.

Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
Results support related syntheses that biomass and taxonomic richness often regrows relatively quickly, 
<!-- **along with traits?,** -->
while taxonomic composition recovers much more slowly, if not diverging altogether *[@norden15]*.
These findings can suggest that more efficient reforestation could more actively target biodiversity conservation over other resilient functions like carbon storage, and that restoration efforts could include focusing activities like native species planting efforts on degraded habitat edges away from primary forest edges, and investing less in areas closer to existing forest edges, where recruitment may already maintain diversity levels.
As forests continue to become fragmented *[@haddad15]*, understanding how to work with natural regeneration patterns around edge effects will likely become increasingly important.

>>>>

This study found significant edge effects on secondary forest stand wood and stem density, tree diversity and community composition, and some functional traits, yet not on overall aboveground biomass.
<!-- These results were shown in Figs \@ref(fig:biomass), \@ref(fig:taxa), and \@ref(fig:traits)... -->
It was initially hypothesized that community composition would vary with distance to high-quality intact primary forest edge, favoring more shade-tolerant taxa, due to light availability and dispersal potential.
The evidence presented here supported edge effects on community composition and biomass of taxa that were associated with both early- and late-successional stages, but without detectable roles for light availability or general dispersal mode.
Accordingly, this suggests that edge effects do significantly change humid tropical forest community taxonomic and functional composition, but primarily due to factors other than light competition or dispersal limitation *[@krishnadas20]*, and also suggests that aboveground biomass and even low taxonomic richness are resilient after a decade of regeneration.

<!-- ## Biomass -->

Aboveground biomass storage is a key global ecosystem function and service, but edge effects tend to release about one-third of carbon lost from tropical deforestation *[@brinck17]*, with more expected in the future *[@mitchard18]*.
<!-- responds variably to secondary succession and edge effects ...   -->
Management has been found to explain most of biomass variation among tropical forest fragments, and wood density ~10% biomass variation *[@pyles22]*.
Additionally, this study contributes that management via edge effects can also mediate ~10% of wood density variation without affecting local biomass, although median stem wood density can describe regional spatial biomass patterns *[@baker04]*.
In this study, wood density and richness values may have been pulled down by early-successional clonal trees closest to the primary forest, which tend to have lower wood densities, lowering carbon *[@gonzalezj98; @pyles22]*, and increasing nearby stem density.
Lower wood density can also reflect more dynamic forest stands *[@malhi06]* with higher tree turnover, a consequence of forest fragmentation *[@nascimento04]* together with other global change drivers *[@laurance14]*, and may also reflect possible differences in soil fertility *[@malhi04]*, all of which represent possible future research directions, beyond strictly neutral dynamics *[@terborgh95]*.
Furthermore, light availability at the plot level did not affect biomass storage, but instead, individual light gaps may be more important for biomass dynamics *[@chazdon84]*.
    <!-- - although canopy light may still affect biomass generally in this case, without edge effects (check direct light-biomass correlation graph? -- no sig correlation) -->

<!-- ## Composition -->

Edge effects have been shown in other studies not only to lower forest biomass, but also to change community composition *[@anderson22]*, and in addition this study contributes that the decline in diversity can be rapid and non-linear across edge gradients, along with supporting the likelihood of changes to community composition lasting decades.
This rapid loss of diversity across the edge gradient may be in part due to existing higher yet constant levels of shade limiting shade-intolerant seedling recruitment and/or performance, together with insufficient time for slower-growing shade-tolerant taxa to accumulate significant amounts of biomass.
Another possible factor may be Janzen-Connell processes *[@wills06; @terborgh20]*, including conspecific negative density-dependence of seedling survival on basal area *[@comita09]*, as well as overall short dispersal kernels, whose effects can be mediated by shade *[@comita14]* and in part by higher fungal pathogen pressure near conspecifics *[@jia20]*.
While both diversity and community composition changed significantly with distance to intact forest in this study, the biomass of most individual taxa, surprisingly, tended to stay the same along the edge gradient.
This lack of significant individual taxon biomass responses along this edge gradient could be due to stochastic population factors affecting their biomass growth, perhaps widening variability.
Regardless of the underlying process, these absent patterns do also help explain the lack of pattern observed with plot biomass overall across the edge gradient.
This explanation points to future studies of taxon-specific responses to edge effects, to help explain overall forest biomass dynamics near habitat edge.
More specifically here, the biomass of the most abundant taxon *Vochysia* was resilient to edge effects in this study, however, that of the taxon *Ficus*, key for frugivores like birds and bats *[@cottee-jones15; @rafidison20]*, was significantly lowered across the secondary forest edge, pointing to potentially different responses of biomass and local food web associated ecosystem services.
Results suggest that this pattern is specific to this locally widespread *Ficus* genus, since the overall biomass of other animal-dispersed taxa as a whole tended to stay the same across the edge gradient studied here.
Interestingly, community composition also appeared to show lower variance across the edge gradient, although testing this observation in a forested area with evenly distributed area among distance strata bands would provide additional support.

<!-- ## Traits -->

Characterizing community composition by functional traits is also increasingly studied *[@kearney21]*, but less so regarding edge effects.
A study in Madagascar found lower phylogenetic diversity closer to degraded forest edge, but no change in aboveground biomass *[@razafindratsima18]*.
This study presents similar results in showing lower taxonomic diversity, although biomass specifically of taxa associated with both early- and late-successional stages also decreased with away from primary forest edge.
This resulting decrease by successional stage could be explained by slightly lower habitat quality inferred near exposed habitat edges near roads, and accordingly relatively higher habitat quality near closed primary forest edge *[@ries04]*.
While lower biomass of succession-agnostic taxa further from primary forest edge appears consistent with the expected forms of group-level dispersal kernels as an explanation, the lack of significant pattern along forest edge when binned by broad dispersal mode that was found here may limit the potential of seed dispersal to explain biomass trends.
Instead, seedling survival may have been more important than dispersal in accumulating basal area and biomass *[@comita09]*.
For taxa primarily dispersed by animals instead of wind, this could be explained by potentially broader local limitations in animal dispersal activity or abundance.
Additionally, the potential for nearby dispersal to facilitate reforestation could be tested in future studies by measuring taxon-specific dispersal kernels.

<!-- ## Conclusion -->

Overall, this study highlights how abandoned wet tropical timber plantation can regenerate in alongside fragmented forests.
Results support related syntheses that tropical forest biomass and taxonomic richness often regrows relatively quickly *[@davies21a]*, 
<!-- **along with traits?,** -->
while taxonomic composition recovers much more slowly, if not diverging altogether *[@norden15]*.
These findings can suggest that more efficient reforestation could more actively target biodiversity conservation over other resilient functions like carbon storage, and that restoration efforts could include focusing activities like native species planting efforts on degraded habitat edges away from primary forest edges, and investing less in areas closer to existing forest edges, where recruitment may already maintain diversity levels.
As forests continue to become fragmented *[@haddad15]*, understanding how to work with natural regeneration patterns around edge effects will likely become increasingly important.

### 2022-11-09 article/intro.Rmd
bioRxiv,poster

Modern secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Effects of degraded edges have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
Additionally, the dominance of pioneering taxa could lead to arrested succession *[@tymen15]*, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
Overall, edge effects ultimately have the potential to shape community composition, in part based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

@@

Influenced by restoration ecology for forest management, this study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, focused on highlighting edge effects from an adjacent primary forest on overall stand properties, tree community composition, and limited functional trait recovery.
<!-- **- what is new in this study?** -->
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
    <!-- - supported by inference?! -->
Accordingly, we predicted more specifically that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

>>>>

Modern secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Effects of degraded edges have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
Additionally, the dominance of pioneering taxa could lead to arrested succession *[@tymen16]*, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
Overall, edge effects ultimately have the potential to shape community composition, in part based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

@@

Influenced by restoration ecology for forest management, this study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, focused on highlighting edge effects from an adjacent primary forest on overall stand properties, tree community composition, and limited functional trait recovery.
<!-- **- what is new in this study?** -->
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
    <!-- - supported by inference?! -->
Accordingly, we predicted more specifically that further from the intact primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

### 2022-11-09 article/merge.Rmd
bioRxiv,poster

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
author:
  - Nicholas Medina * ^a^
  # - Emma Capetz ^c^
  - Andrea Johnson ^b^
  # - Erin Lau ^c^
  # - Agustín Mendoza ^b^
  # - Don Morris ^c^
  - Max Villalobos ^b^
date: \scriptsize
  ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA;
  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
  # ^c^ Earth and Environmental Sciences, Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
  # ^a^ Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

output: 
  # html_notebook
  # bookdown::html_document2:
  #   # df_print: paged
  #   global_numbering: TRUE
  #   toc: FALSE
  bookdown::pdf_document2:
    # number_sections: FALSE
    global_numbering: TRUE
    keep_tex: TRUE
    toc: FALSE
bibliography: 
  - ../../library.bib
  - ../../grateful-refs.bib
  - ../../software.bib
link-citations: TRUE
csl: ../../styles-master/elsevier-harvard.csl

fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespace
# abstract
---

@@

***Corresponding author**: Nicholas Medina, nmedina@umich.edu

@@

This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship.

# Declaration of interests{-}

@@

Thanks to 
Professor Dan Perlman for initial networking;
<!-- Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;  -->
concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
James Fifer, Chau Ho, and Michelle Spicer for discussion of early drafts;
and Luisa Valdez for translation assistance.

@@

MV and AJ contributed to study conceptualization, administration, and supervision;
NM and AJ contributed to funding acquisition;
NM contributed to data collection;
NM and MV contributed to data analysis;
NM wrote initial draft and later revisions.

>>>>

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a reforested wet neotropical timber plantation"
author:
  - Nicholas Medina * ^a,d^
  - Emma Capetz ^c^
  - Andrea Johnson ^b^
  - Agustín Mendoza ^b^
  - Max Villalobos ^b^
date: \scriptsize
  ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA;
  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Puntarenas, Costa Rica;
  ^c^ Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
  ^d^ Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
# author:
#   - name: Nicholas Medina
#     email: nmedina@umich.edu
#     affliation: 
#       - Brandeis University
#       - University of Michigan
#     # orcid_id: 0000-0001-5465-3988
#     correspondingauthor: TRUE
#   - name: Emma Capetz
#     affiliation: Lehigh University
#     email: emmacapetz11@gmail.com
#   - name: Andrea Johnson
#     affiliation: Osa Conservation
#     email: andrea.e.johnson@gmail.com
#   # - name: Erin Lau
#   #   affiliation: Lehigh University
#   #   # email:
#   - name: Agustín Mendoza
#     affiliation: Osa Conservation
#     email: agustinescaladorosa@gmail.com
#   - name: Max Villalobos
#     affiliation: Osa Conservation
#     email: maxvillalobosv@gmail.com
# address:
#   - code: Brandeis University
#     address: 415 South St, Waltham, MA 02453 USA
#   - code: Osa Conservation
#     address: Puerto Jiménez, Puntarenas, Costa Rica
#   - code: Lehigh University
#     address: 27 Memorial Drive West, Bethelhem, PA 18015 USA
#   - code: University of Michigan
#     address: 1105 N University Ave, Ann Arbor, MI 48109 USA
# keywords:
#   - secondary forest
#   - regeneration
#   - timber plantation associated biodiversity
#   - edge effect
#   - aboveground biomass
#   - shade  
# abstract: |
#   About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
#   Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
#   Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
#   However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of nearby degraded forests.
#   Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
#   This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
#   It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
#   A wet neotropical timber plantation reforested after 20 years and adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched identified taxa to broad dispersal mode and wood density traits using relevant literature.
#   As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light surprisingly tended to stay the same.
#   Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, explained in part by wood density, and taxonomic composition varied notably. 
#   Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
#   Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.
#   
#   Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
#   Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
#   Dado que la mayoría de bosques quedan entre un km de su borde, efectos de borde probablemente afectan la regeneración de composición de comunidad.
#   Sin embargo, la mayoría de estudios evalúan cómo bordes ya degradados afectan a bosques intactos, dejándolo poco claro si bordes de más alta calidad podrían facilitar la regeneración de bosques cercanos previamente degradados.
#   Notablemente, bordes de alta calidad cerca de bosques intactos podrían promover procesos como disperso y acumulación de biomasa maderosa que efectivamente acceleran a la sucesión, permitiendo mejor rendimiento de taxa tolerantes a sombra comparado con taxa más adaptados a alta luz en las etapas tempranas de regeneración de plantaciones foresterados.
#   Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
#   Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.
#   Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10% variación explicado, mientras tanto biomasa y luz de dócel asombrosamente tendió a quedarse igual.
#   Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, explicado en parte por densidad de madera, y composición taxonomica varió notablemente.
#   Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
#   Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

output: 
  # html_notebook
  # bookdown::html_document2:
  #   # df_print: paged
  #   global_numbering: TRUE
  #   toc: FALSE
  bookdown::pdf_document2:
    number_sections: FALSE
    global_numbering: TRUE
    keep_tex: TRUE
    toc: FALSE
  # bookdown::pdf_book:
    # base_format: rticles::springer_article
    # citation_package: biblatex
bibliography: 
  - ../../library.bib
  - ../../grateful-refs.bib
  - ../../software.bib
link-citations: TRUE
csl: ../../styles-master/elsevier-harvard.csl

fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{setspace}\doublespace
---

@@

***Correspondence**: Nicholas Medina, nmedina@umich.edu

@@

This study was funded by Brandeis University Hiatt Career Center World of Work Fellowship and in part by University of Michigan Rackham Merit Fellowship.

<!-- # Declaration of interests{-} -->

<!-- Authors declare no conflicts of interest.  -->

Thanks to 
Brandeis University Professor Dan Perlman for initial networking;
Lehigh University Professor Don Morris 
and student Erin Lau 
for field support;
concurrent Osa Conservation general manager Manuel Ramírez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;
James Fifer, Chau Ho, and Michelle Spicer for review of early draft content;
and Luisa Valdez for translation assistance.

@@

@@

NM, MV and AJ contributed to study conceptualization, design, administration, and supervision;
NM and AJ contributed to funding acquisition;
NM, AM, and EC, 
contributed to key data collection;
NM and MV contributed to data analysis;
NM wrote initial draft and later revisions.

Study team includes and recognizes diverse and historically-excluded contributions to research; with bilingual communication efforts completed and pending.
